---
title: "mrwater Assumptions"
author: "Felicitas D. Beier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrwater Assumptions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, echo = TRUE, eval = FALSE, comment = "#>")
```

```{r Settings, echo = FALSE}
library(mrwater)
library(stringr)

lpjml <- c(natveg = "LPJmL4_for_MAgPIE_44ac93de",
           crop = "ggcmi_phase3_nchecks_bft_e511ac58")
climatetype <- "GSWP3-W5E5:historical"
```

## Overview

The following document provides an overview of assumptions that
the mrwater library is based on. 

### Treatment of data mismatches

## Zero Irrigation Water Requirements

```{r readirrigWatReqVSirrigAreas,  include = FALSE, echo = FALSE}
# Irrigation water requirements per cell per crop
# given irrigation system (in m^3 per hectare per year)
cropIrrigReq <- calcOutput("ActualIrrigWatRequirements",
  irrigationsystem = "initialization",
  selectyears = "y2010", iniyear = "y2010",
  lpjml = lpjml, climatetype = climatetype,
  multicropping = FALSE, aggregate = FALSE)
cropIrrigReq <- collapseNames(cropIrrigReq[, , "consumption"])

# physical croparea
croparea <- calcOutput("CropareaAdjusted",
  iniyear = iniyear,
  aggregate = FALSE)

irArea <- collapseNames(croparea[, , "irrigated"])
rfArea <- collapseNames(croparea[, , "rainfed"])
```

The allocation of potentially irrigated areas (PIA) is based on the total volume of
water requirements (sum of water requirements over all crops grown under the selected cropmix)
that is compared to the available water (see ``calcRiverDischargeAllocation``).
For the case that crop-specific irrigation water requirements are reported to be 0 in the 
input data, no irrigation area is assigned to these crops in the algorithm that derives PIA
because the assumption is that irrigation is not required there and therefore rainfed production
takes place (see ``calcIrrigAreaPotential``). 
```{r RainfedLandMismatch,  include = TRUE, echo = TRUE}
print(paste0("Total rainfed area is, ",
              round(sum(rfArea), digits = 2),
              " Mha."))
print(paste0("Of these, ",
             round(sum(rfArea[cropIrrigReq == 0]), digits = 2),
             " Mha have 0 water requirements."))
print(paste0("That is, ",
             round(sum(rfArea[cropIrrigReq == 0]) / sum(rfArea) * 100, digits = 2),
             " % of rainfed land."))
```

However, for the scenario of 'committed agricultural uses' where areas that are already irrigated
are reserved for irrigation before the allocation of additional potentials, areas that are reported
to be irrigated but show 0 irrigation water requirements (ca. 0.43 Mha, i.e. 0.18% of irrigated area),
areas are assigned to be irrigated (see ``calcIrrigAreaActuallyCommitted``) to avoid data mismatches
in the area accounting.
```{r IrrigatedLandMismatch,  include = TRUE, echo = TRUE}
print(paste0("Total irrigated area is, ",
             round(sum(irArea), digits = 2),
             " Mha."))
print(paste0("Of these, ",
             round(sum(irArea[cropIrrigReq == 0]), digits = 2),
             " Mha have 0 water requirements."))
print(paste0("That is, ",
             round(sum(irArea[cropIrrigReq == 0]) / sum(irArea) * 100, digits = 2),
             " % of irrigated land."))
```

