---
title: "mrwater Documentation"
author: "Felicitas D. Beier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{mrwater Documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r Setup, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, echo = TRUE, eval = FALSE, comment = "#>")
```

## Overview

The following document provides a full code documentation of the ``mrwater``
library including a range of arguments that can be varied to produce cellular
pre-processed water-related outputs.
The ``fullWATER.R`` gives an overview of functions generating the main model
outputs. It can be executed with the function call ``madrat:::retrieveData(WATER)``
that will generate a .tgz file with all relevant outputs.


### Related libraries

The ``mrwater`` library is part of the mr-universe. For more details to the
related libraries, please have a look at the respective libraries:

```{r Libraries}
library(madrat)
library(magclass)
library(mrcommons)
library(mrland)
```

### Input data

The ``mrwater`` library relies on a range of input data. Open-source data is
directly downloaded and imported via ``downloadSource`` and ``readSource``
functions that are part of the ``madrat`` library.
Data that is not openly available needs to be stored in a specific folder
structure (``inputdata/sources/DataSource``) where the respective DataSource
folder must have the same name as the ``readSource`` function
(e.g. readRamankutty refers to the inputdata/sources/Ramankutty folder).
The read functions of this library and the underlying libraries necessary to
run the code (e.g. ``mrwater:::readIrrigationSystem``; ``mrcommons:::readFAO``;
``mrland:::readZabel2014``) access this folder to include the required data.


## Motivation

The mrwater library can be used to generate stand-alone spatially explicit
irrigation potentials (irrigation water potentials (withdrawals and consumption)
and irrigation area potentials) and (aggregated) input data for land-system
models. Potential aggregation units are spatial clusters (based on bio-physical
similarity of grid cells), country- or basin-scale.

Furthermore, its outputs are a useful disaggregation tool for land-system model
outputs. For example irrigation water withdrawals from land-use models such as
MAgPIE can be disaggregated to a 0.5-degree spatial resolution using spatially
explicit irrigation potentials.

It is designed as hydrological input data processing tool for the global land-use
model MAgPIE (Model of Agricultural Production and its Impact on the Environment).
Nevertheless, it is applicable for other land-system models and its spatially
explicit as well as aggregated outputs can be used for a variety of
land-system models (e.g. CGEs, PEs, etc.).

Note: To return spatially explicit function outputs, the ``madrat:::calcOutput()``
function call must include the argument ``aggregate = FALSE``. For aggregation,
a mapping is required.


## Hydrological input data

In this version of the mrwater library, all hydrological input data is provided
by the process-based hydrology-vegetation model LPJmL.
Data that is based on natural vegetation runs is provided by LPJmL4. Data that
includes management is based on LPJmL5.

The river routing consists of several functions.
The first iteration determines naturalized discharge based on yearly runoff and
lake evaporation using the STN river structure
(``rs <- readRDS(system.file("extdata/riverstructure_stn_coord.rds", package = "mrwater"))``)
that determines the flow direction and basin attribution of the 67420 grid cells
of the underlying land mask.
The function returns natural discharge and lake evaporation that are required
for further calculations in the river routing routine.

```{r NatDischarge}
natQ <- calcOutput("RiverNaturalFlows", selectyears = 2010,
                    climatetype = "GFDL-ESM4:ssp126",
                    lpjml = c(natveg = "LPJmL4_for_MAgPIE_44ac93de",
                    crop = "ggcmi_phase3_nchecks_9ca735cb"), aggregate = FALSE)[, , "discharge_nat"]
```

The required arguments are ``selectyears`` (determining the year(s) for which
the output shall be generated), ``climatetype`` (selecting the GCM data and
scenario underlying the process-based vegetation model), and ``lpjml``
(selecting the LPJmL model versions used for natural vegetation (``natveg``)
input data and management (``crop``) input data)

Environmental flow requirements (EFRs) and accessible discharge are also
calculated based on hydrological LPJmL data only. Both are based on monthly discharge.
The calculation method and strictness of EFR can be selected via the ``efrMethod``
argument (e.g. "Smakhtin:good", "VMF:fair").
Inaccessible discharge is based on the assumption that the higher the long-term
seasonal variability of discharge is, the harder it is to access the water by
humans and bring it into productive use. Different accessibility rules can be
selected via the ``accessibilityrule`` argument (e.g. "CV:2" stands for the
coefficient of variation approach to a base of 2; "Q:75" stands for a quantile
approach where discharge that exceeds the 75th quantile is inaccessible to humans).

```{r EFRs}
selectyears <- 2010
lpjml       <- c(natveg = "LPJmL4_for_MAgPIE_44ac93de", crop = "ggcmi_phase3_nchecks_bft_6277d36e")
climatetype <- "GFDL-ESM4:ssp126"

efr <- calcOutput("EnvmtlFlowRequirements", efrMethod = "VMF:fair",
                  selectyears = selectyears, climatetype = climatetype, lpjml = lpjml,
                  aggregate = FALSE)[, , "EFR"]

inaccessibleQ <- calcOutput("DischargeInaccessible", accessibilityrule = "CV:2",
                            selectyears = selectyears, climatetype = climatetype, lpjml = lpjml,
                            aggregate = FALSE)

efrMethod         <- "VMF:fair"
accessibilityrule <- "CV:2"
```

Note: The integration of alternative biophysical and hydrological input data is
possible.


## Human water use data

Human water abstractions are considered in consecutive river routing iterations
that reserve the respective water abstraction and update cellular discharge
respectively. They are determined through the following function call:

```{r HumanDischarge}
humanQ <- calcOutput("RiverDischargeNatAndHuman", iniyear = 2010, comAg = TRUE,
                     selectyears = selectyears, multicropping = FALSE,
                     climatetype = climatetype, lpjml = lpjml,
                     efrMethod = efrMethod, aggregate = FALSE)

iniyear <- 2010
comAg   <- TRUE
```

Non-agricultural water abstractions as provided by ISIMIP (and for future scenarios WATERGAP) have priority over agricultural water abstractions. 
They are determined through the function call 
``calcOutput("RiverHumanUses", humanuse = "non_agriculture", ...)`` 
and report reserved non-agricultural water consumption ("currHuman_wc") and withdrawal ("currHuman_ww").

Taking so-called "committed agricultural uses", i.e. reserving water abstractions
of currently irrigated areas, into account is optional (it can be activated via the ``com_ag`` argument).
In case of an activation of committed agricultural uses, the required consumption and withdrawal for currently irrigated areas as determined in
``calcOutput("WaterUseCommittedAg", ...)`` are reserved in addition to 
non-agricultural water abstractions via the function call 
``calcOutput("RiverHumanUses", humanuse = "committed_agriculture", ..., iniyear = 2010)``.
The argument ``iniyear`` determines the chosen cellular irrigated crop mix 
(i.e. the crop mix as reported by FAOSTAT for the year 2010 on cellular cropland as reported by LUH2). 
Cellular irrigated areas for specific years are provided via 
``calcOutput("Croparea", years = iniyear, sectoral = "kcr", physical = TRUE, cells = "lpjcell", cellular = TRUE, irrigation = TRUE, aggregate = FALSE)``
through combining (irrigated and rainfed) cropland area provided by LUH2 and country-level FAO crop production data. Note that areas reported as irrigated depreciate at a rate of 10 percent per annum for future time steps (years after ``iniyear``).

Compared to the previous section, this section requires additional input data:
ISIMIP non-agricultural water abstraction data (for the historical time period 
and fixed at this level for the future), WATERGAP non-agricultural water 
abstraction data (for future scenarios),  LPJmL5 crop water requirements, 
country level irrigation system shares as provided by Jägermeyr (2015), 
LUH2 cellular cropland area (see also: ``mrcommons:::calcLUH2v2``), 
and FAO country-level crop-specific production data (see also: ``mrcommons:::readFAO_online``).


## Land-use input data

Spatial-explicit land-use and irrigation area data provided by LUH2
(see also ``mrcommons:::calcLUH2v2``).
Since the LUH2 cropland map is subdivided into only five crop functional types
(C3 annuals; C4 annuals; C3 perennials; C4 perennials; C3 nitrogen fixers),
we use the spatial distribution of LUH2 and FAOSTAT information on total
country-level crop-specific production to derive spatially-explicit crop-specific
areas (see ``mrcommons:::calcCroparea``). The country-level FAOSTAT cropmix is
distributed equally across physical cropland of the respective country while
maintaining the spatially explicit cropland area distribution as provided by LUH2.
The distribution of rice area is derived more explicitly. LUH2 provides cellular
flooded area shares and only rice is flooded according to the data set.
The distribution of physical rice areas is therefore determined by assigning
the country's rice production to flooded areas provided at cellular level by LUH2.
Aerobic (non-paddy) rice is accounted by distributing country-level FAO rice areas 
beyond country-aggregated LUH2 flooded area (i.e. where FAO reports higher 
country-level rice areas than there are LUH2 flooded areas in the respective country) 
equally across the remaining country's cropland area.
Note that flooded areas are not accounted as irrigated areas.
For one, because flooded rice production is often only partially irrigated
and also because it fulfills a special management purpose in terms of pest control.

```{r Landuse}
croparea <- calcOutput("Croparea", years = iniyear, sectoral = "kcr",
physical = TRUE, cells = "lpjcell", cellular = TRUE, irrigation = TRUE, aggregate = FALSE)
```

The ``physical`` argument determines accounts for the cropping intensity in that
 ``physical = TRUE`` returns cropland areas that match physical area.
 With ``physical = FALSE`` croparea can exceed physical land area, due to
 multicropping.


## Irrigation Water Potentials (IWP)

Finally, the function ``calcRiverSurplusDischargeAllocation`` determines the
allocation of "surplus discharge", i.e. the discharge of the estuary cell that
is not (yet) consumed along the river in the last iteration of the previous
river routings that reserved environmental and human water uses. The function
returns the water that is potentially available for irrigation considering
biophysical, economic and management constraints.

```{r AllocationAlgorithm}
iwp <- calcOutput("RiverDischargeAllocation", output = "potIrrigWat",
                   selectyears = selectyears, climatetype = climatetype, lpjml = lpjml,
                   efrMethod = efrMethod, accessibilityrule = accessibilityrule,
                   iniyear = iniyear, com_ag = com_ag,
                   allocationrule = "optimization",
                   rankmethod = "USD_ha:TRUE", yieldcalib = TRUE, cropmix = "hist_total",
                   thresholdtype = "USD_ha", gainthreshold = 500,
                   landScen = "potCropland:HalfEarth",
                   irrigationsystem = "initialization", multicropping = FALSE, aggregate = FALSE)
```


### Alternative allocation algortihms

The ``mrwater`` library includes different surplus discharge allocation rules
representing different management strategies ("upstreamfirst" vs. "optimization")
that can be set in the ``allocationrule`` argument.
The upstreamfirst algorithm allows irrigation water abstractions provided that
there is sufficient local discharge available and a positive yield gain through
irrigation reserving these irrigation water abstractions in upstream grid cells
first following the calculation order of cells from upstream to downstream
(for more details on the river structure and calculation order see ``riverstructure.Rmd``).
The optimization algorithm ranks grid cells according to their yield value gain
through irrigation allocating available local discharge to the grid cell with
the highest productivity first, such that water resources are made available
at the most efficient location rather than upstream before downstream.


### Economic yield value gains

The yield value gain can be expressed in USD per hectare irrigated area or USD
per cubic meter of irrigation water set via the ``unit`` argument
in ``calcIrrigYieldImprovementPotential`` and via the ``rankmethod`` argument
in ``calcRiverSurplusDischargeAllocation``. The ``rankmethod`` argument
consists of two components: the unit selections ("USD_ha" or "USD_m3") and a
boolean determining whether grid cells are ranked according to their full yield
gain potential (TRUE) or taking a two step approach ranking according to their
75% potential and their full potential allocating 50% of irrigation water
requirements in each of the two iterations (FALSE) separated by a ":".

The yield gain through irrigation - as calculated from LPJmL5 irrigated and
rainfed yields - is valued with global average crop prices as provided by
FAOSTAT (see ``mrcommons:::calcOutput("IniFoodPrice", datasource = "FAO",
products = "kcr", years = NULL, year = iniyear, aggregate = FALSE))``.

Depending on the modeling application, LPJmL irrigated and rainfed crop yields
can be calibrated to meet FAO crop yields via the ``yieldcalib`` argument.
The ``cropmix`` argument determines the selection of crops ("hist_total" refers
to the cellular cropmix as determined in ``calcCroparea`` considering irrigated
and rainfed areas, "hist_irrig" refers to cellular cropmix as determined in
``calcCroparea`` considering only irrigated areas. Besides these crop mixes,
a selection of one or several proxycrop(s) can be selected. This setting is
relevant when calculating irrigation potentials for the future).

```{r YieldValueGain}
yieldGain <- calcOutput("IrrigYieldImprovementPotential", unit = "USD_ha:GLO",
                        selectyears = selectyears,
                        climatetype = climatetype, lpjml = lpjml,
                        iniyear = iniyear,
                        cropmix = "hist_total", yieldcalib = TRUE,
                        multicropping = FALSE, aggregate = FALSE)
```

To represent economic aspects of irrigation, different yield value gain
thresholds can be chosen to calculate the respective IWP. It is set via
the arguments ``thresholdtype`` and ``gainthreshold``.
The chosen valuation method is determined via the unit argument that determines
whether water or irrigation area is valued (USD_ha or USD_m3) and whether 
global average prices (GLO) or country-level agricultural prices (ISO) are used.

### Work-in-progress

Please refrain from using the ``multicropping = TRUE`` argument.
This implementation is still work-in-progress and will be updated soon.


### Cropland extent

The irrigation potentials can be calculated for various scenarios of potential
cropland cover that are set via the ``landScen`` argument.
It consists of two parts: (1) the potentially available land component where
currently irrigated area (currIrrig), current cropland area (currCropland) or
potential cropland area (potCropland) can be selected, and (2) a protection
component that selects which land areas shall be protected, i.e. no irrigation
taking place in these areas. Possible selections are:
WDPA, BH, FF, CPD, LW, BH_IFL, HalfEarth.
For example ``landScen = "currCropland:NA"`` selects current cropland extent as
of the year set in iniyear without land protection;
``landScen = "potIrrig:HalfEarth"``
is potential cropland while respecting a Half Earth land protection scenario.


### Irrigation Area Potential (IAP)

The ``calcIrrigatableArea`` function translates potential irrigation water (PIW)
into potentially irrigated areas (PIA) and returns the area that can potentially
be irrigated given the assumptions that are set via the arguments.

```{r IAP}
allocationrule   <- "optimization"
rankmethod       <- "USD_ha:TRUE"
yieldcalib       <- TRUE
cropmix          <- "hist_total"
thresholdtype    <- "USD_ha"
gainthreshold    <- 500
landScen         <- "potCropland:HalfEarth"
irrigationsystem <- "initialization"
multicropping    <- FALSE

iap <- calcOutput("IrrigAreaPotential", selectyears = selectyears, iniyear = iniyear,
                   lpjml = lpjml, climatetype = climatetype,
                   efrMethod = efrMethod, accessibilityrule = accessibilityrule,
                   rankmethod = rankmethod, yieldcalib = yieldcalib, allocationrule = allocationrule,
                   thresholdtype = thresholdtype, gainthreshold = gainthreshold,
                   irrigationsystem = irrigationsystem, landScen = landScen,
                   cropmix = cropmix, com_ag = com_ag,
                   potential_wat = TRUE, multicropping = FALSE, aggregate = FALSE)
```


### Generation of potential irrigation area (PIA) curves

For a representation of irrigation potentials at various irrigation yield value
gains, the river routing routine has to be executed for varying irrigation yield
value gain thresholds.
The generated data can be used to create PIA curves and/or aggregated to different
scales (e.g. country-, basin-scale).

```{r EconOfIrrig}
iad <- calcOutput("EconOfIrrig", scenario = "ssp2", output = "IrrigArea",
             gtrange = c(0, 250, 500, 1000, 2000, 3000), selectyears = 2010, iniyear = 2010,
             lpjml = lpjml, climatetype = climatetype, efrMethod = efrMethod, accessibilityrule = accessibilityrule,
             rankmethod = rankmethod, yieldcalib = yieldcalib, transDist = transDist,
             allocationrule = allocationrule, thresholdtype = thresholdtype,
             irrigationsystem = irrigationsystem, landScen = landScen, cropmix = cropmix,
             potential_wat = TRUE, com_ag = FALSE, multicropping = FALSE, aggregate = FALSE)
```
